Type theory in type theory with single substitutions

### 2024-09-15 types_2024_postproc/p.tex
stuff

@@

@@

>>>>

We conjecture that the same works for any SOGAT.

@@

In this section we show that if have we universes and $\Pi$ types
(which is the case in the theory (\ref{eq:tt})), we can decrease the
number of euqations.

In our single substitution calculus, we have the equation $[\p][^+]$
stated both for types and terms (we know that the left one is about
types because of our convention that type metavariables are uppercase
Latin letters and term metavariables are lowercase):
\[
B[\p][\gamma^+] = B[\gamma][\p] \hspace{4em} b[\p][\gamma^+] =_{[\p][^+]} b[\gamma][\p]
\]
We even need the first equation to typecheck the second one. We made
the dependency explicit by adding a subscript of the equality in the
equation for terms (in Agda, this dependency has to be made
explicit). An alternative presentation of the second equation without
requiring the first one is the following:
\[
[\p][^+]' : (e : B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_{e} b[\gamma][\p]
\]
That is, for any type $B$ for which $B[\p][\gamma^+] = B[\gamma][\p]$,
we have that for any $b$ we have the equation (suitably over the input
equation). Thus we add a conditional equation.

It turns out that if we have Coquand-universes, the conditional
$[\p][^+]'$ rule is enough: the input equation holds for $B := \U\,i$
via $\U[]$, thus we get that for any $\hat{A} : \Tm\,\Gamma\,(\U\,i)$,
$\hat{A}[\p][\gamma^+] = \hat{A}[\gamma][\p]$. But every type has a
code in the universe, so for any $B : \Ty\,\Gamma\,i$,
\begin{alignat*}{10}
  & B[\p][\gamma^+] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p][\gamma^+] && {=}(\El[]) \\
  & \El\,((\cd\,B)[\p][\gamma^+])\,\, && {=}([\p][^+]') \\
  & \El\,((\cd\,B)[\gamma][\p]) && {=}(\El[]) \\
  & \El\,(\cd\,B)[\gamma][\p] && {=}(\U\beta) \\
  & B[\gamma][\p].
\end{alignat*}
Equation $[\q][^+] : \q[\gamma^+] = \q$ only makes sense if we have
$[\p][^+]$ for types, so either we make $[\q][^+]$ conditional, or we
coerce it along the derived equation.

We play the exact same game with $[\p][\langle\rangle]$: we replace it
with the conditional equation
\[
[\p][\langle\rangle]' : (e:B[\p][\langle a\rangle] = B)\ra b[\p][\langle a\rangle] =_e b,
\]
notice that we have the input for $B = \U\,i$, and derive the input
equation for all types.

Now we turn our attention to the equation
\[
  [\langle\rangle][] : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle].
\]
We were forced to introduce it so that we can state the substitution
rule for function application $\cdot[]$. What if we made $\cdot[]$
itself conditional?

xxx

@@

\pagebreak

\appendix

\section{The single substitution GAT presentation of the SOGAT (\ref{eq:tt})}

For reference, we list the rules of the theory introduced in Section
\ref{sec:ssc}. The core substitution calculus:
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
& \Sub && : \Con\ra\Con\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
& [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
& [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
& \q[^+] && : \q[\gamma^+] = \q \\
& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
& \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
& [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B
\end{alignat*}
Rules for individual type formers.
\begin{alignat*}{10}
  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
  & \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \\
  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
  & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
  & \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
  & \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
